---
title: "project_wrangling"
output:
  html_document:
    df_print: paged
date: "2026-10-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, escape = FALSE, warning = FALSE, message = FALSE, fig.width = 5, fig.height = 3)
```

```{r}

# install tidyverse and simple features if not installed
# Note: sf requires RTools

packages = c("tidyverse", "sf")

for (package in packages){
  if ( !(package %in% installed.packages()) ){
    install.packages(package)
    }
}

library(tidyverse)
library(sf)

```

## **Dataset:**

**Remind us which dataset(s) you are exploring.**

We will explore the Community Tree Priority Map. According to their webpage, the map enables Austin organizations and individuals to decide where activities like planting, tree care, and community outreach could occur around Austin.

**Download the dataset from the City of Austin Data Portal as a *.csv* file and import it into R.**

```{r}
url = "https://data.austintexas.gov/api/v3/views/ugty-5b95/query.csv"

# if my_trees csv does not exist, download it from url
if(!file.exists("my_trees.csv")){
  download.file(url, "my_trees.csv")
}

# if my_trees obj doesn't exist, import it using read
if (!exists(x = "my_trees")){
  my_trees = read.csv("my_trees.csv")
}

# removing first 4 col that came from the API
colnames(my_trees)
my_trees_vars = my_trees |>
  select(-c(X.id, X.version, X.created_at, X.updated_at))
```

**Report the number of rows and columns in the dataset.**

```{r}
# reports as rows then columns
dim(my_trees)

```

**Show the first few rows and the main variables of interest.**

```{r}
# The core vars of interest are: zip codes, service area, along with each of the 4 individual scores: environment, social, investment and health.
colnames(my_trees)

# other variables that may be interesting are the_geometry & priority_score
my_trees_vars = my_trees |>
  select(
    # region related
    zipcodes,
    ae_service_area,
    # the four individual scores
    environment_score,
    social_score,
    investment_score,
    health_score,
  )

# first 5 rows of my_trees_vars
head(my_trees_vars, n = 5)
```

**If you are using more than one dataset, explain how you would join the datasets.**

At this time we are not planning on using more than this one dataset.

## **Research Question:**

**Remind us of your research question**

How do environmental, social, investment and health scores vary across ZIP codes or within/outside service areas and which urban factors most strongly correlate with a high tree priority score?

## **Variables**

#### Variable `zipcodes`

**What it represents**

`zipcodes` represents the postal codes that the the geometry of the tract is in. One tract may partially overlap with multiple postal codes.

**How we will treat it**

We will treat `zipcodes` as a nominal categorical variable

**Visualizations**

```{r}
# ggplot of single categorical
my_trees_vars |>
  count(zipcodes) |>
  arrange(desc(n)) |>
  slice_head(n = 15) |>
  ggplot(aes(x = n, y = reorder(zipcodes, n))) +
  geom_col(fill = "lightblue") +
  labs(x = "Count", y = "Zipcodes") +
  theme_minimal()
```

**Summary Statistics**

```{r}
# summary statistics for categorical
my_trees_vars |>
  separate_rows(zipcodes, sep = ",") |>
  count(zipcodes, name = "frequency") |>
  mutate(proportion = frequency / sum(frequency))

```

------------------------------------------------------------------------

#### Variable `ae_service_area`

**What it represents**

`ae_service_area` represents whether the census tract is completely within, intersects, or is outside of the Austin Energy service area. The Austin Energy service area can be used as an approximate border of the City of Austin.

**How we will treat it**

We will treat `ae_service_area` as a nominal categorical variable.

**Visualizations**

```{r}
# ggplot of single categorical
my_trees_vars |>
  count(ae_service_area) |>
  arrange(desc(n)) |>
  ggplot(aes(x = n, y = ae_service_area)) +
  geom_col(fill = "lightblue") +
  labs(
    x = "Count", y = "Service Area Proximity"
  ) +
  theme_minimal()
```

**Summary Statistics**

```{r}
# summary statistics for categorical
my_trees_vars |>
  count(ae_service_area, name = "frequency") |>
  mutate(proportion = frequency/sum(frequency))
```

------------------------------------------------------------------------

#### Variable `environment_score`

**What it represents**

`environment_score` represents the cumulative score for the "Natural Environment" category. It is calculated by normalizing three factors: tree canopy coverage, flood plain coverage, and mean maximum land surface temperature.

**How we will treat it**

We will treat `environment_score` as a continuous numeric variable.

**Visualizations**

```{r}
# ggplot of single continuous numeric variable
ggplot(data = my_trees_vars) +
  aes(x = environment_score) +
  geom_histogram(fill = "lightgreen", binwidth = 4) +
  theme_minimal() +
  labs(x = "Environment Score", y = "Frequency")
```

**Summary Statistics**

```{r}
# summary statistics for a continuous numeric variable
summary(my_trees_vars$environment_score)
```

------------------------------------------------------------------------

#### Variable `social_score`

**What it represents**

`social_score` represents the cumulative score for the "Social Vulnerability" category. It is taken from the social vulnerability index which is from the UT Austin "Austin Area Sustainability Indicators".

**How we will treat it**

We will treat `social_score` as a continuous numeric variable

**Visualizations**

```{r}
# ggplot of single continuous numeric
ggplot(data = my_trees_vars) +
  aes(x = social_score) +
  geom_histogram(fill = "darkgoldenrod1", binwidth = 5) +
  theme_minimal() +
  labs(x = "Social Score", y = "Frequency")
```

**Summary Statistics**

```{r}
# summary statistics for continuous numeric
summary(my_trees_vars$social_score)
```

------------------------------------------------------------------------

#### Variable `investment_score`

**What it represents**

`investment_score` represents the cumulative score for the "Community Investment" category. It is calculated by normalizing two factors: urban forest funding by neighborhood and tree plantings from NeighborWoods, ReadySetPlant and Austin community tree programs per square mile.

**How we will treat it**

We will treat `investment_score` as a continuous numeric variable

**Visualizations**

```{r}
# ggplot of single continuous numeric
ggplot(data = my_trees_vars) +
  aes(x = investment_score) +
  geom_histogram(fill = "paleturquoise", binwidth = 5) +
  theme_minimal() +
  labs(x = "Investment Score", y = "Frequency")
```

**Summary Statistics**

```{r}
# summary statistics for continuous numeric
summary(my_trees_vars$investment_score)
```

------------------------------------------------------------------------

#### Variable `health_score`

**What it represents**

`health_score` represents the cumulative score for the "Health and Well-Being" category. It is calculated by normalizing three factors: self-reported physical and mental health data from the CDC and projected roadway emissions for 2035 as a stand-in for air quality.

**How we will treat it**

We will treat `health_score` as a continuous numeric variable

**Visualizations**

```{r}
# ggplot of single continuous numeric
ggplot(data = my_trees_vars) +
  aes(x = health_score) +
  geom_histogram(fill = "firebrick1", binwidth = 5) +
  theme_minimal() +
  labs(x = "Health Score", y = "Frequency")
```

**Summary Statistics**

```{r}
# summary statistics for continuous numeric
summary(my_trees_vars$health_score)
```

------------------------------------------------------------------------

#### Aggregate Visualizations and Summaries

**Preparing our Scores**

```{r}
# scores pivoted
scores = my_trees_vars |>
  select(colnames(my_trees_vars)[grep("score", x = colnames(my_trees_vars))]) |>
  pivot_longer(
    everything(),
    names_to = "Type",
    values_to = "Score"
    )
```

**Summary Statistics**

```{r}
# summary statistics of score, note extreme difference in variance
scores |>
    group_by(Type) |>
    summarize(
        mean = mean(Score),
        median = median(Score),
        sd = sd(Score),
        minimum = min(Score),
        maximum = max(Score),
        )
```

**Plot**

```{r}
# faceted histogram of scores
ggplot(data = scores, aes(x = Score, fill = Type)) +
    geom_histogram(binwidth = 10) +
    facet_wrap(~Type) +
    theme(legend.position = "none")
```

**Interpretation**

Social score has a very high variance. With such a high variance, social score would explain much of the variance in the overall priority score. Investment score mostly takes on values near 100 and has the least variance.

## **Wrangling:**

#### **Individual Changes**

First, we removed the columns that came from the API such as X.id, X.version, etc.

```{r}
# removing rows that came from the API
my_trees = my_trees |>
  select(-c(X.id, X.version, X.created_at, X.updated_at))

```

We further narrowed the dataset to our variables of interest (these may change)

```{r}
my_trees_vars = my_trees |>
  select(
    # region related
    zipcodes,
    ae_service_area,
    # the four individual scores
    environment_score,
    social_score,
    investment_score,
    health_score,
  )

```

For one exploration of our data we pivoted the data frame to focus on how scores change

```{r}
# individual tree scores pivoted
scores = my_trees_vars |>
  select(colnames(my_trees_vars)[grep("score", x = colnames(my_trees_vars))]) |>
  pivot_longer(
    everything(),
    names_to = "Type",
    values_to = "Score"
    )
```

And for another made a single vector of zipcodes to see how frequently they occurred. The original had some rows with multiple zip codes. Now each index is a single zip code.

```{r}
my_trees_vars |>
  separate_rows(zipcodes, sep = ",") |>
  count(zipcodes, name = "frequency") |>
  mutate(proportion = frequency / sum(frequency))

```

Our data was already very well-organized and ready for the kind of analysis we want to do so there was minimal wrangling required. We had no NA values, etc.

#### **Summary**

**What changes did you make to the original data?**

Primarily we removed data we were not interested in, separated out the zip codes, and made a new data frame to further investigate the variation in the individual categories of tree priority scores.

**Which variables will you use in your analysis?**

Presently we are using zip codes, service area, and the four numeric scores: environment_score, health_score, investment_score and social score.

**Is there anything else you think may need to be addressed before beginning your exploratory data analysis?**

We would like to pursue using the "the_geometry" column which contains GeoJSON formatted data using the Simple Features package. That requires some further investigation into how the package works, and may require some tidying of that column. If we do use it, it may replace zip codes for location information.

## **Notes**

<https://www.treeequityscore.org/analyzer/austin/map> - The new tree equity system, interactive. Can be used to show how the two systems score the same areas. Interesting!

<https://auf.isa-arbor.com/content/early/2025/07/16/jauf.2025.020> - Table 5 in this article shows how the individual scores (health, investment, etc.) are calculated and their data sources. This table was the source for how individual scores were calculated.

We may want to use `the_geometry` since it contains coordinates that can be used alongside a single line transecting Austin. We can then define East/West divides or North/South divides, etc just by applying a rule there. If so, we may want to reformat it into separate columns for ease of use.
